---
title-block-banner: true
title: "Exploratory data analysis for NeuroColombia research project"
subtitle: "Data analysis notebook"
date: today
date-format: full
author: 
  - name: "Daniel Manrique-Castano"
    orcid: 0000-0002-1912-1764
    degrees:
      - PhD
    affiliation: 
      - name: Digital Research Alliance of Canada 
        department: Research Data Management
note: "GitHub: https://daniel-manrique.github.io/"
keywords: 
  - Epidemiology
  - Neuropsychology
  - Mental healh
  - Colombia 
   
license: "CC BY"

format:
   typst:
    toc: true
    toc-depth: 2
    columns: 1
   html:
    code-fold: false
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    theme: spacelab

knitr:
  opts_chunk: 
    warning: false
    message: false
---

## Install and load packages

We install the required packages to perform exploratory data visualization and statistical inference. 

```{r}

if (!requireNamespace("brms", quietly = TRUE)) install.packages("brms")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("ggridges", quietly = TRUE)) install.packages("ggridges")
if (!requireNamespace("ggbeeswarm", quietly = TRUE)) install.packages("ggbeeswarm")
if (!requireNamespace("gt", quietly = TRUE)) install.packages("gt")
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
if (!requireNamespace("hrbrthemes", quietly = TRUE)) install.packages("hrbrthemes")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")


library(brms)
library(dplyr)
library(ggplot2)
library(ggridges)
library(ggbeeswarm)
library(gt)
library(here)
library(hrbrthemes)
library(tidyr)


```

We load additional functions

```{r}
Plot_theme <- theme_classic() +
  theme(
      plot.title = element_text(size=18, face="bold", hjust = 0.5, vjust = 2),
      plot.subtitle = element_text(size = 10, color = "black"),
      plot.caption = element_text(size = 12, color = "black"),
      axis.line = element_line(colour = "black", linewidth = 1.5, linetype = "solid"),
      axis.ticks.length=unit(7,"pt"),
     
      axis.title.x = element_text(colour = "black", size = 16, vjust = -2),
      axis.text.x = element_text(colour = "black", size = 16, angle = 0, hjust = 0.5),
      axis.ticks.x = element_line(colour = "black", linewidth = 1),
      
      axis.title.y = element_text(colour = "black", size = 16, vjust = 3),
      axis.text.y = element_text(colour = "black", size = 16),
      axis.ticks.y = element_line(colour = "black", linewidth = 1),
      
      legend.position="right",
      legend.direction="vertical",
      legend.title = element_text(colour="black", face="bold", size=12),
      legend.text = element_text(colour="black", size=10),
      
      plot.margin = margin(t = 10,  # Top margin
                             r = 5,  # Right margin
                             b = 10,  # Bottom margin
                             l = 10) # Left margin
      ) 
```


# Load the dataset

```{r}
Neuro_Data <- read.csv("Data_Processed/Ministery_DiagnosticData_Compiled.csv")
```


# Temporal evolution of neuropsychological diagnostics

In the first place, we will expero the temporal evoluction of neuropsychological diagnostics in Colombia. 


# Total number of cases

First, we create a summary of the number of cases to show the overall trend.

```{r}
Summary_year<- Neuro_Data %>%
  group_by(Year) %>%
  summarise(Total_Diagnostics = sum(Cases, na.rm = TRUE)) %>%
  ungroup()

Summary_year
```

We create a line visualizatyion using `ggplot`:

```{r}
#| label: Diagnostics_year-line
#| fig-cap: Distribution of neuropsychological diagnostics per year
#| fig-width: 6
#| fig-height: 5

Diagnostic_Year_fig <- Summary_year %>%
  ggplot( aes(x= Year, y=Total_Diagnostics)) +
    geom_line( color="grey", size= 2) +
    geom_point(shape=21, color="black", fill="#69b3a2", size=6) +
    ggtitle("Neuropsychological diagnostics") +
  labs(y = "Number of diagnostics",
       x = "Year") +
    Plot_theme

ggsave(Diagnostic_Year_fig,
       filename = "Plots/Diagnostic_Year_fig.png",
       width    = 15, 
       height   = 10, 
       units    = "cm")

Diagnostic_Year_fig 
```

Looking at the distribution in @Diagnostics_year-line, we can see that 2020 strongly disturbs the trend because of the COVID-19 pandemic, which limited the number of diagnostics in the whole country. For this reason, we consider it relevant in further analysis to make a separate analysis for the years before and after COVID. Therefore, we create the required auxiliary variables for the main `Neuro_Data` dataset:
 
```{r}
Neuro_Data <- Neuro_Data %>%
  mutate(Period = case_when(
    Year >= 2016 & Year <= 2019 ~ "Pre-Covid",
    Year >= 2020 & Year <= 2022 ~ "Post-Covid",
    TRUE ~ NA_character_
  ))
```

## Distribution per year and capital city 

We group the data to obtain the cases per year, Capital and sex

```{r}
Summary_Year_Capital <- Neuro_Data %>%
  group_by(Year, Department, Capital) %>%
  summarise(Total_Cases = sum(Cases, na.rm = TRUE),
            Total_Population = sum(Population, na.rm = TRUE)) %>%
  ungroup()

Summary_Year_Capital$Percentage <- (Summary_Year_Capital$Total_Cases / Summary_Year_Capital$Total_Population) * 100

Summary_Year_Capital |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE)
```

After visualizing the data, we plot it as a scatterplot.

```{r}
#| label: Diagnostics_year-scatter
#| fig-cap: Distribution of neuropsychological diagnostics per year
#| fig-width: 6
#| fig-height: 5

Year_Capital_fig <- Summary_Year_Capital %>%
  ggplot(aes(x=Year, y=Percentage, color = Capital)) +
    geom_jitter(shape=21, size=1) +
    ggtitle("Neuropsychological diagnostics") +
  labs(y = "Percentage of diagnostics",
       x = "Year") +
  Plot_theme +
  theme (legend.position = "bottom",
         legend.direction = "horizontal") 

ggsave(Year_Capital_fig,
       filename = "Plots/Year_Capital_fig.png",
       width    = 15, 
       height   = 15, 
       units    = "cm")

Year_Capital_fig
```
Overall, @Diagnostics_year-scatter shows that the percentage of diagnoses is higher in non-capital cities. This is an aspect that we can model further.


# Distribution per diagnostic

Now, we visualize the distribution of diagnostics.

```{r}

Summary_Diagnostic <- Neuro_Data %>%
  group_by(Diagnostic) %>%
  summarise(Total_Cases = sum(Cases, na.rm = TRUE)) %>%
  ungroup()

Summary_Year_Capital |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE
  )

```

We pint a bar plot using the `geom_bar` function:

```{r}
#| label: Diagnostics_Diagnostic
#| fig-cap: Distribution of neuropsychological diagnostics per diagnostic type (CIE-11)
#| fig-width: 6
#| fig-height: 10


# Create a bar plot with bars ordered from highest to lowest
Diagnostic_fig <- ggplot(Summary_Diagnostic, 
                         aes(y = reorder(Diagnostic, -Total_Cases), 
                             x = Total_Cases,
                             fill = Total_Cases)) +
  geom_bar(stat = "identity") +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  geom_text(aes(label=Total_Cases), vjust=0.5, color="black", size=4)+
  labs(x = "Total cases", y = "Diagnostic", title = "Number of cases per diagnostic") +
  Plot_theme +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "none")
  

ggsave(Diagnostic_fig,
       filename = "Plots/Diagnostic_fig.png",
       width    = 16, 
       height   = 30, 
       units    = "cm")

Diagnostic_fig 
```

Based on @Diagnostics_Diagnostic, we subset and write a table keeping the main diagnostics for further analysis.

```{r}
Neuro_Data_MainDiag <- Neuro_Data  %>%
  filter(Diagnostic %in% c("F900", "F809", "F808", "F801", "F819", "F800", "F818", "F813", "F840"))

write.csv(Neuro_Data_MainDiag, "Data_Processed/Neuro_Data_MainDiag.csv", row.names = FALSE)
```


## Distribution per diagnostic and sex

Now, we are interestied in exploring the distribution of diagnostics by sex

```{r}
Summary_Diagnostic_Sex <- Neuro_Data_MainDiag %>%
  group_by(Department, Diagnostic, Sex) %>%
  summarise(Total_Cases = sum(Cases, na.rm = TRUE),
            Total_Population = sum(Population, na.rm = TRUE)) %>%
  ungroup()

Summary_Diagnostic_Sex$Percentage <- (Summary_Diagnostic_Sex$Total_Cases / Summary_Diagnostic_Sex$Total_Population) * 100

Summary_Diagnostic_Sex |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE
  )
```
We visualize the data for the main diagnostics as density plots using `ridges`

```{r}
#| label: Diagostics_sex
#| fig-cap: Distribution of neuropsychological diagnostics per biological sex
#| fig-width: 6
#| fig-height: 5

Diagnostic_Sex_fig <- Summary_Diagnostic_Sex %>%
  ggplot(aes(x=Percentage, y=Diagnostic, color = Sex, fill = Sex)) +
    geom_density_ridges2(alpha = 0.4) +
    ggtitle("Neuropsychological diagnostics") +
    xlim(0, 1) +
  labs(y = "Diagnostic",
       x = "Percentage of diagnostics") +
  Plot_theme +
  theme (legend.position = "bottom",
         legend.direction = "horizontal") 

ggsave(Diagnostic_Sex_fig,
       filename = "Plots/Diagnostic_Sex_fig.png",
       width    = 15, 
       height   = 15, 
       units    = "cm")

Diagnostic_Sex_fig
```

@Diagostics_sex shows that men are more likely to recieve some diagnostics.

## Cases by sex and age

Now, we include age as a variable in the distribution by sex. 

```{r}
Summary_Age_Sex <- Neuro_Data_MainDiag %>%
  group_by(Age, Sex) %>%
  summarise(Total_Cases = sum(Cases, na.rm = TRUE),
            Total_Population = sum(Population, na.rm = TRUE)) %>%
  ungroup()

Summary_Age_Sex$Percentage <- (Summary_Age_Sex$Total_Cases / Summary_Age_Sex$Total_Population) * 100

Summary_Age_Sex |>
  gt() |>
  opt_interactive(
    use_search = TRUE,
    use_filters = TRUE,
    use_resizers = TRUE
  )
```


```{r}
#| label: Diagostics_sex-age
#| fig-cap: Distribution of neuropsychological diagnostics per biological sex and age
#| fig-width: 6
#| fig-height: 5

Age_Sex_fig <- Summary_Age_Sex %>%
  ggplot(aes(x= Age, y= Total_Cases, color = Sex)) +
    geom_line(size= 2) +
    geom_point(shape=21, color = "black", fill = "black", size=6) +
    ggtitle("Neuropsychological diagnostics") +
  scale_x_continuous(breaks = seq(0, 11, by = 2)) +
  labs(y = "Number of diagnostics",
       x = "Age") +
  Plot_theme +
  theme (legend.position = "bottom",
         legend.direction = "horizontal") 

ggsave(Age_Sex_fig ,
       filename = "Plots/Age_Sex_fig .png",
       width    = 15, 
       height   = 15, 
       units    = "cm")

Age_Sex_fig
```
@Diagostics_sex-age shows that males are more likely to be diagnosed, although both sexes peak at 6 years.Given the change in the trend, we find suitable to model this behaviours using splines. 
